---
title: "R Plots Implemented with Base R and ggplot2"
author: "DSA400W Class at Chatham University, Fall 2021"
date: "9/21/2021"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This document will explore the implementation of various types of plots that can be produced with base R graphics and/or the ggplot2 package. For each plot type we begin by showing basic plots and then enhance the plots by combining plot types, adding new variables using faceting, color, shape or size, or customizing colors, axes, axis labels, annotations or styles.

## Histogram

## Density Curve

## Bar Plot

### Bar plot with data added above the bars (Base R)
First, let's read in some data on passengers on the *Titanic*.

```{r}
titanic <- read.csv("data/titanic.csv")
```

Next, we calculate survival percentage and class count for each passenger
class, using the `aggregate()` function, and then merge those two results into a data frame.
```{r}
# calculate survival percentage by class and count by class.
surv_pct <- aggregate(Survived ~ Pclass, data = titanic, FUN = mean)
class_ct <- aggregate(Survived ~ Pclass, data = titanic, FUN = length)

# Merge those two results into a single data frame, and specify 
# column names
by_class_data <- merge(surv_pct, class_ct,
                       by = "Pclass")
colnames(by_class_data) <- c("Pclass", "Surv_Pct", "Count")
#str(by_class_data)
```

Now create the barplot. Note that we increase the `ylim` parameter a bit to leave room for labels above the bars. We also set `axes = FALSE` so that the default y-axis is not plotted. Note also that we assign the barplot object to a variable. What gets assigned to the variable is the x positions of the midpoints of each of the bars on the plot. We will use those x positions later to position the labels above the bars, since the x position for a text label specifies the center of the label.

In the next step we create a custom y-axis with the `axis()` function. The `at` parameter specifies the placement of the tick marks on the axis ("axis ticks") and the `labels` parameter specifies the tick labels. Use the `paste()` function to make percentage labels.

Finally, we create the labels showing the count of passengers in each class and place them right above the bars. The x position comes from the barplot object and the y position is just the height of the bars plus a little extra so that the text labels don't overlap the bars.

```{r}
plt <- barplot(by_class_data$Surv_Pct, 
               names.arg = paste("Class", by_class_data$Pclass),
               col = 'steelblue',
               main = "Survival Percentage by Passenger Class",
               ylab = "Survival Percentage",
               xlab = "Passenger Class",
               ylim = c(0.0, 0.75),
               axes = FALSE
)

# Add y axis
axis(2, at = seq(0.0, 0.7, by = 0.1),
     labels = paste(seq(0, 70, by = 10), "%", sep=""),
     las = 1,
     cex.axis = 0.9)

# Add the text of the labels
text(x = plt, 
     y = by_class_data$Surv_Pct + 0.04, 
     labels = paste("n: ", by_class_data$Count, sep=""), 
     cex=1) 
```

## Line Plot ("Time Series Plot")

## Stacked Area Plot

## Boxplot

## Violin Plot

## Strip Plot

## Normal Probability (QQ) Plot

## Mosaic Plot

## Scatterplots

## Bubble Plot

## Diverging Bar Chart

## Diverging Lollipop Chart

## Chloropleth




