---
title: "R Plots Implemented with Base R and ggplot2"
author: "DSA400W Class at Chatham University, Fall 2021"
date: "9/21/2021"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Overview

This document will explore the implementation of various types of plots that can be produced with base R graphics and/or the ggplot2 package. For each plot type we begin by showing basic plots and then enhance the plots by combining plot types, adding new variables using faceting, color, shape or size, or customizing colors, axes, axis labels, annotations or styles.

## Histogram

Based on example Histogram from:
https://statisticsglobe.com/histogram-in-base-r-hist-function

Using the rivers dataset from Base R
```{r}
rivers
```

### Basic Histogram (Base R)

```{r}
hist(rivers)
```

Adding titles, changing the color of the histogram, changing the breaks, changing the axis limits

```{r}
hist(rivers,                                                # Change main title of histogram
     main = "Length of Rivers in North America", col='blue', breaks=20, xlim=c(0, 5000), ylim=c(0, 120))
```

Using the Iris dataset to customize a histogram.

```{r}
iris
```

Displaying the default histogram for petal width.

```{r}
hist(iris$Petal.Width)
```

Customizing the histogram.

```{r}
hist(iris$Petal.Width,                                                
     main = "Iris Petal Width Histogram", xlab = " Petal Width", col='pink', breaks=20, xlim=c(0, 3), ylim=c(0, 50))
```

Re-creating a histogram with ggplot2 based on the following example:
http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization

Creating the dataset.

```{r}
set.seed(1234)
df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5), rnorm(200, mean=65, sd=5)))
  )
head(df)
```

Creating a basic histogram.

```{r}
# Basic histogram
ggplot(df, aes(x=weight)) + geom_histogram()
# Change the width of bins
ggplot(df, aes(x=weight)) + 
  geom_histogram(binwidth=1)
# Change colors
p<-ggplot(df, aes(x=weight)) + 
  geom_histogram(color="black", fill="white")
p
```

Creating a histogram for Iris Dataset on Petal Length using ggplot2 and customizing.

```{r}
##Basic histogram
ggplot(iris, aes(x=iris$Petal.Length)) + geom_histogram()
##Changing the bins
ggplot(iris, aes(x=iris$Petal.Length)) + 
  geom_histogram(binwidth=1.5)
##Changing the colors and adding a title
s<-ggplot(iris, aes(x=iris$Petal.Length)) + 
  geom_histogram(color="yellow", fill="red", binwidth=1.5)
s + ggtitle("Iris Petal Length Histogram") + xlab("Petal Length")
```

## Density Curve

### Basic Density Curve Using Base R
```{r}
# get a dataset from R
# Orange dataset has information about the growth of orange trees
orange <- Orange
head(orange)
```

```{r}
# create a density plot
plot(density(orange$circumference),
     main = "CIRCUMFERENCE OF ORANGE TREES",
     xlab = "CIRCUMFERENCE",
     ylab = "DISTRIBUTION")

polygon(density(orange$circumference), col=7, border=2)
```

### Basic Density Curve Using GGPlot2
```{r}
# create the density plot

orange %>%
  ggplot( aes(x = circumference)) +
  geom_density(fill= 7, color= 2)

```
## Bar Plot
###Base R
```{r}
### Create a data set to work with
survey <- c(apple=45, orange=25, honeydew=35, banana=50, watermelon=20, blueberry=15)

survey

### Create a barchart with titles and color
barplot(survey,
        main="Favorite Fruit Survey",
        xlab="Fruit",
        ylab="Count",
        col=c("red", "orange", "green", "yellow", "pink", "blue"))

### Add a legend
legend("topright",
       legend = c("apple", "orange", "honeydew", "banana", "watermelon", "blueberry"),
       col = c("red", "orange", "green", "yellow", "pink", "blue"),
       pch = 19,
       bty = "n",
       pt.cex = 2,
       cex = 1.2)

```

### GGplot2
```{r}
# First load ggplot2
library(ggplot2)

# Use the same data set created in the base r example but make it into a data frame
survey1 <- data.frame(
  fruit=c("apple", "orange", "honeydew", "banana", "watermelon", "blueberry"),
  value=c(45, 25, 35, 50, 20, 15))

# Create the barchart
ggplot(survey1, aes(x=fruit, y=value)) + geom_bar(stat="identity") 

# Add color with black outlines
ggplot(survey1, aes(x=fruit, y=value)) + geom_bar(stat="identity", fill=c("red", "orange", "green", "yellow", "pink", "blue"), color="black") + labs(x="Count", y="Fruit", title="Fruit Survey") + theme(legend.position = "none")

```
### Bar plot with data added above the bars (Base R)
First, let's read in some data on passengers who were aboard the *Titanic*.

```{r}
titanic <- read.csv("data/titanic.csv")
```

Next, calculate survival percentage and class count for each passenger
class, using the `aggregate()` function, and then merge those two results into a data frame.
```{r}
# calculate survival percentage by class and count by class.
surv_pct <- aggregate(Survived ~ Pclass, data = titanic, FUN = mean)
class_ct <- aggregate(Survived ~ Pclass, data = titanic, FUN = length)

# Merge those two results into a single data frame, and specify 
# column names
by_class_data <- merge(surv_pct, class_ct,
                       by = "Pclass")
colnames(by_class_data) <- c("Pclass", "Surv_Pct", "Count")
#str(by_class_data)
```

Create the barplot. Note that we increase the `ylim` parameter a bit to leave room for labels above the bars. We also set `axes = FALSE` so that the default y-axis is not plotted. Note also that we assign the barplot object to a variable. What gets assigned to the variable is the x positions of the midpoints of each of the bars on the plot. We will use those x positions later to position the labels above the bars, since the x position for a text label specifies the center of the label.

In the next step we create a custom y-axis with the `axis()` function. The `at` parameter specifies the placement of the tick marks on the axis ("axis ticks") and the `labels` parameter specifies the tick labels. Use the `paste()` function to make percentage labels.

Finally, we create the labels showing the count of passengers in each class and place them right above the bars. The x position comes from the barplot object and the y position is just the height of the bars plus a little extra so that the text labels don't overlap the bars.

```{r}
plt <- barplot(by_class_data$Surv_Pct, 
               names.arg = paste("Class", by_class_data$Pclass),
               col = 'steelblue',
               main = "Survival Percentage by Passenger Class",
               ylab = "Survival Percentage",
               xlab = "Passenger Class",
               ylim = c(0.0, 0.75),
               axes = FALSE
)

# Add y axis
axis(2, at = seq(0.0, 0.7, by = 0.1),
     labels = paste(seq(0, 70, by = 10), "%", sep=""),
     las = 1,
     cex.axis = 0.9)

# Add the text of the labels
text(x = plt, 
     y = by_class_data$Surv_Pct + 0.04, 
     labels = paste("n: ", by_class_data$Count, sep=""), 
     cex=1) 
```

### Bar plot with data added above the bars (ggplot2)

First, let's read in some data on passengers who were aboard the *Titanic*. For the `ggplot2` example we will use the `tidyverse` approach.

```{r message = FALSE, results = "hide"}
titanic <- read_csv("data/titanic.csv")
```

Next, calculate survival percentage and class count for each passenger
class.

```{r}
survival <- titanic %>%  
  group_by(Pclass) %>% 
  summarize(Count = n(),
            Surv_Pct = mean(Survived))
```

Make the plot.

```{r}
ggplot(survival, aes(x = Pclass, y = Surv_Pct)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste("n: ", Count, sep="")), vjust = -0.5) +
  xlab("Passenger Class") +
  ylab("Survival Percentage") +
  ggtitle("Survival Percentage by Passenger Class") +
  scale_y_continuous(limits = c(0, 0.70),
                     breaks = seq(0, .7, by = 0.10),
                     labels = paste(seq(0, 70, by = 10), "%", sep=""))  +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) # Centers title
```

## Line Plot ("Time Series Plot")

## Stacked Area Plot

## Boxplot
In the Boxplot section add two basic boxplot examples, one utilizing base R and one utilizing ggplot2. Note that even basic examples should have appopriate titles and axis labels.

Base R:
```{r}
# Some data to work with:
data(ChickWeight)
```
```{r}
head(ChickWeight)
str(ChickWeight)
```
```{r}
#Boxplot:
boxplot(ChickWeight$weight ~ ChickWeight$Time,
        main = "Chick Age versus Chick Weight",
        xlab = "Chick Age in Days",
        ylab = "Chick Weight in Grams")
```
ggplot2:
```{r}
library(ggplot2)
p <- ggplot(ChickWeight, aes(group=Time, y=weight)) + 
  geom_boxplot() +
  labs(title="Chick Age versus Chick Weight",x="Chick Age in Days", y = " Chick Weight in Grams")+
  theme_classic()
p
```

## Violin Plot

## Strip Plot

## Normal Probability (QQ) Plot

A quantile-quantile ("QQ") plot is used to compare an empirical distribution to a theoretical distribution by plotting the observed quantiles vs the theoretical quantiles. They are typically used to evaluate whether or not a variable is normally distributed.

The built-in `rivers` dataset gives the lengths (in miles) of 141 rivers in North America, as compiled by the US Geological Survey. Let's see if the river lengths are normally distributed by creating a normal probability plot of the river lengths.

### Basic QQ Plot with Normal Distributoin Reference Line (Base R)

Here we create a basic QQ plot in base R and add the normal distribution QQ plot as a reference line for comparison.
```{r}
qqnorm(rivers,
       main = "Q-Q Plot of North American River Lengths")
qqline(rivers, col = "red")
```

### Basic QQ Plot with Normal Distributoin Reference Line (ggplot2)

Here we produce a similar plot using ggplot2. First we need to put the river lengths in a data frame.

```{r}
rivers_df <- data.frame(length = rivers)
```

Now we make the plot. Note that the shape and size of the geom are revised to better match the base R plot.

```{r}
ggplot(rivers_df, aes(sample = length)) +
  stat_qq(shape = 1, cex = 2) +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of North American River Lengths",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) # Centers title
```


## Mosaic Plot

## Scatterplots

## Bubble Plot

### Base R Bubble Plot

Base R plots require a bit finessing to get just right. 

First of all, I plan to place the legend outside the plot itself so I must first change the margins. This is done with `par()` which can set graphical parameters. Within `par()`, `mar` takes a vector input of 4 numbers determining how far into the pane the plot margins should be, starting at the bottom and continuing clockwise. `xpd` takes a logical value & setting it to TRUE tells R not to clip the legend or any other insertion outside the plot bounds. `inset=c(-0.35,0)` tells R to move the legend away from the plot, outside the margin horizontally. To make the legend look a bit cleaner, use `bty='n'` to remove the box around the legend. 

To set the point size in base plot, use `cex=data$variable`. In this case `cex=mtcars$disp`. For the legend, use `pt.cex=c(n1, n2, n3...nx)` for however many legend items you have. I'm choosing intervals of 100 from 100 to 400. 

```{r}
par(mar=c(4, 5, 2, 8), xpd = T) # bottom, left, top, right

plot(mtcars$mpg, mtcars$hp, cex = mtcars$disp,
     pch=16,
     main = 'Miles per Gallon vs Gross Horsepower with Displacement',
     xlab = 'miles per gallon',
     ylab = 'gross horspower')
legend('right', inset=c(-0.35,0),
       legend=c('100', '200', '300', '400'),
       title = 'Displacement\n(cu. in.)',
       pch=16, ncol=1, pt.cex = c(100, 200, 300, 400),
       bty = 'n')
```

Yikes, it seems like the points may be a bit too large. Lets make them all smaller by dividing their values by 100 & take care not to forget about the legend points as well by changing the values for `pt.cex`. 

```{r}
par(mar=c(4, 5, 2, 8), xpd = T) # bottom, left, top, right

plot(mtcars$mpg, mtcars$hp, cex = mtcars$disp/100,
     pch=16,
     main = 'Miles per Gallon vs Gross Horsepower with Displacement',
     xlab = 'miles per gallon',
     ylab = 'gross horspower')
legend('right', inset=c(-0.35,0),
       legend=c('100', '200', '300', '400'),
       title = 'Displacement\n(cu. in.)',
       pch=16, ncol=1, pt.cex = c(100, 200, 300, 400)/100,
       bty = 'n')
```

Better, but the points don't seem to be proportional. It seems that `cex` scales points by altering the radius of a point. This would cause the area of each consecutive point to be exponentially larger than the last when using the formula $A=πr^2$. To correct this and make the point size proportional to the area, rearrange the formula to solve for the radius that produces the desired area. $r=\sqrt{\frac{A}{π}}$

Now the points are proportional in area but still too large, so I divide the areas by 5. 

```{r}
par(mar=c(4, 5, 2, 8), xpd = T) # bottom, left, top, right

plot(mtcars$mpg, mtcars$hp, cex = sqrt(mtcars$disp/3.14)/5,
     pch=16,
     main = 'Miles per Gallon vs Gross Horsepower with Displacement',
     xlab = 'miles per gallon',
     ylab = 'gross horspower')
legend('right', inset=c(-0.35,0),
       legend=c('100', '200', '300', '400'),
       title = 'Displacement\n(cu. in.)',
       pch=16, ncol=1, pt.cex = sqrt(c(100, 200, 300, 400)/3.14)/5,
       bty = 'n')
```

Now it looks right!

### Ggplot2 Bubble Plot

When I need to make a plot, I usually prefer `ggplot()` over base r for its simplicity. As you can see, I was able to make a clean graph with just 5 lines of code. To show  displacement as a variable point size, you set `size=disp` within the aesthetics of ggplot. Ggplot automatically scales by area & makes the points reasonable sizes as well. Because mtcars is set as the data source, there is no need to reference the data frame for each variable. 

```{r}
ggplot(mtcars, aes(x=mpg, y=hp, size=disp))+
  geom_point()+
  ggtitle('Miles per Gallon vs Gross Horsepower with Displacement')+
  labs(size = 'Displacement\n(cu. in.)', x='miles per gallon', y='gross horsepower')+
  theme_classic()
```

## Diverging Bar Chart
```{r}
data(swiss)
head(swiss)
str(swiss)
```
```{r}

#Data Preparation
library(ggplot2)
swiss$`municipality` <- rownames(swiss)  # create new column for municipality names
swiss$Fertility_z <- round((swiss$Fertility - mean(swiss$Fertility))/sd(swiss$Fertility), 2)  # compute normalized mpg
swiss$Fertility_type <- ifelse(swiss$Fertility_z < 0, "below", "above")  # above / below avg flag
swiss <- swiss[order(swiss$Fertility_z), ]  # sort
swiss$`municipality` <- factor(swiss$`municipality`, levels = swiss$`municipality`)  # convert to factor to retain sorted order in plot.

#Diverging Barplot:
ggplot(swiss, aes(x=`municipality`, y=Fertility_z, label=Fertility_z)) + 
  geom_bar(stat='identity', aes(fill=Fertility_type), width=.5) +
  theme(text = element_text(size=9)) +
  scale_fill_manual(name="Fertility", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="#00ba38", "below"="#f8766d")) + 
  labs(title= "Divergence in the Average Fertility in Swiss Municipalities",
       x = "Municipality",
       y = 'Fertility') + 
  coord_flip()

```
## Diverging Lollipop Chart

## Chloropleth




