---
title: "R Plots Implemented with Base R and ggplot2"
author: "DSA400W Class at Chatham University, Fall 2021"
date: "9/21/2021"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Overview

This document will explore the implementation of various types of plots that can be produced with base R graphics and/or the ggplot2 package. For each plot type we begin by showing basic plots and then enhance the plots by combining plot types, adding new variables using faceting, color, shape or size, or customizing colors, axes, axis labels, annotations or styles.

## Histogram

Based on example Histogram from:
https://statisticsglobe.com/histogram-in-base-r-hist-function

Using the rivers dataset from Base R
```{r}
rivers
```

### Basic Histogram (Base R)

```{r}
hist(rivers)
```

Adding titles, changing the color of the histogram, changing the breaks, changing the axis limits

```{r}
hist(rivers,                                                # Change main title of histogram
     main = "Length of Rivers in North America", col='blue', breaks=20, xlim=c(0, 5000), ylim=c(0, 120))
```

Using the Iris dataset to customize a histogram.

```{r}
iris
```

Displaying the default histogram for petal width.

```{r}
hist(iris$Petal.Width)
```

Customizing the histogram.

```{r}
hist(iris$Petal.Width,                                                
     main = "Iris Petal Width Histogram", xlab = " Petal Width", col='pink', breaks=20, xlim=c(0, 3), ylim=c(0, 50))
```

Re-creating a histogram with ggplot2 based on the following example:
http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization

Creating the dataset.

```{r}
set.seed(1234)
df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5), rnorm(200, mean=65, sd=5)))
  )
head(df)
```

Creating a basic histogram.

```{r}
# Basic histogram
ggplot(df, aes(x=weight)) + geom_histogram()
# Change the width of bins
ggplot(df, aes(x=weight)) + 
  geom_histogram(binwidth=1)
# Change colors
p<-ggplot(df, aes(x=weight)) + 
  geom_histogram(color="black", fill="white")
p
```

Creating a histogram for Iris Dataset on Petal Length using ggplot2 and customizing.

```{r}
##Basic histogram
ggplot(iris, aes(x=iris$Petal.Length)) + geom_histogram()
##Changing the bins
ggplot(iris, aes(x=iris$Petal.Length)) + 
  geom_histogram(binwidth=1.5)
##Changing the colors and adding a title
s<-ggplot(iris, aes(x=iris$Petal.Length)) + 
  geom_histogram(color="yellow", fill="red", binwidth=1.5)
s + ggtitle("Iris Petal Length Histogram") + xlab("Petal Length")
```

## Density Curve

### Basic Density Curve Using Base R
```{r}
# get a dataset from R
# Orange dataset has information about the growth of orange trees
orange <- Orange
head(orange)
```

```{r}
# create a density plot
plot(density(orange$circumference),
     main = "CIRCUMFERENCE OF ORANGE TREES",
     xlab = "CIRCUMFERENCE",
     ylab = "DISTRIBUTION")

polygon(density(orange$circumference), col=7, border=2)
```

### Basic Density Curve Using GGPlot2
```{r}
# create the density plot

orange %>%
  ggplot( aes(x = circumference)) +
  geom_density(fill= 7, color= 2)

```
## Bar Plot

### Bar plot with data added above the bars (Base R)
First, let's read in some data on passengers who were aboard the *Titanic*.

```{r}
titanic <- read.csv("data/titanic.csv")
```

Next, calculate survival percentage and class count for each passenger
class, using the `aggregate()` function, and then merge those two results into a data frame.
```{r}
# calculate survival percentage by class and count by class.
surv_pct <- aggregate(Survived ~ Pclass, data = titanic, FUN = mean)
class_ct <- aggregate(Survived ~ Pclass, data = titanic, FUN = length)

# Merge those two results into a single data frame, and specify 
# column names
by_class_data <- merge(surv_pct, class_ct,
                       by = "Pclass")
colnames(by_class_data) <- c("Pclass", "Surv_Pct", "Count")
#str(by_class_data)
```

Create the barplot. Note that we increase the `ylim` parameter a bit to leave room for labels above the bars. We also set `axes = FALSE` so that the default y-axis is not plotted. Note also that we assign the barplot object to a variable. What gets assigned to the variable is the x positions of the midpoints of each of the bars on the plot. We will use those x positions later to position the labels above the bars, since the x position for a text label specifies the center of the label.

In the next step we create a custom y-axis with the `axis()` function. The `at` parameter specifies the placement of the tick marks on the axis ("axis ticks") and the `labels` parameter specifies the tick labels. Use the `paste()` function to make percentage labels.

Finally, we create the labels showing the count of passengers in each class and place them right above the bars. The x position comes from the barplot object and the y position is just the height of the bars plus a little extra so that the text labels don't overlap the bars.

```{r}
plt <- barplot(by_class_data$Surv_Pct, 
               names.arg = paste("Class", by_class_data$Pclass),
               col = 'steelblue',
               main = "Survival Percentage by Passenger Class",
               ylab = "Survival Percentage",
               xlab = "Passenger Class",
               ylim = c(0.0, 0.75),
               axes = FALSE
)

# Add y axis
axis(2, at = seq(0.0, 0.7, by = 0.1),
     labels = paste(seq(0, 70, by = 10), "%", sep=""),
     las = 1,
     cex.axis = 0.9)

# Add the text of the labels
text(x = plt, 
     y = by_class_data$Surv_Pct + 0.04, 
     labels = paste("n: ", by_class_data$Count, sep=""), 
     cex=1) 
```

### Bar plot with data added above the bars (ggplot2)

First, let's read in some data on passengers who were aboard the *Titanic*. For the `ggplot2` example we will use the `tidyverse` approach.

```{r message = FALSE, results = "hide"}
titanic <- read_csv("data/titanic.csv")
```

Next, calculate survival percentage and class count for each passenger
class.

```{r}
survival <- titanic %>%  
  group_by(Pclass) %>% 
  summarize(Count = n(),
            Surv_Pct = mean(Survived))
```

Make the plot.

```{r}
ggplot(survival, aes(x = Pclass, y = Surv_Pct)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste("n: ", Count, sep="")), vjust = -0.5) +
  xlab("Passenger Class") +
  ylab("Survival Percentage") +
  ggtitle("Survival Percentage by Passenger Class") +
  scale_y_continuous(limits = c(0, 0.70),
                     breaks = seq(0, .7, by = 0.10),
                     labels = paste(seq(0, 70, by = 10), "%", sep=""))  +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) # Centers title
```

## Line Plot ("Time Series Plot")

## Stacked Area Plot

## Boxplot

## Violin Plot

## Strip Plot

## Normal Probability (QQ) Plot

## Mosaic Plot

## Scatterplots

## Bubble Plot - Ian

```{r}
ggplot(mtcars, aes(x=mpg, y=hp, size=disp))+
  geom_point()+
  ggtitle('Miles per Gallon vs Gross Horsepower with Displacement')+
  labs(size = 'Displacement\n(cu. in.)', x='miles per gallon', y='gross horsepower')+
  theme_classic()
```

```{r}
par(mar=c(4, 5, 2, 8), xpd = T) # bottom, left, top, right

plot(mtcars$mpg, mtcars$hp, cex = sqrt(mtcars$disp/3.14)/5,
     pch=16,
     main = 'Miles per Gallon vs Gross Horsepower with Displacement',
     xlab = 'miles per gallon',
     ylab = 'gross horspower')
legend('right', inset=c(-0.35,0),
       legend=c('100', '200', '300', '400'),
       title = 'Displacement\n(cu. in.)',
       pch=16, ncol=1, pt.cex = sqrt(c(100, 200, 300, 400)/3.14)/5,
       bty = 'n')
```


## Diverging Bar Chart

## Diverging Lollipop Chart

## Chloropleth




